---
title: "Text Preprocessing"
format: revealjs
editor: visual
---

```{r}
#| include: false
exam_data <- readRDS("../Data/01_raw/data_raw.rds")
```

## Load Libraries

```{r}
#| include: false
# Load packages
library(dplyr)
library(stringr)
library(textclean)
library(stringdist)
library(stopwords)
library(spacyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(ggalluvial)
library(xtable)
```

## Filter Data

This study only includes three subjects: biology, economics, and nature and chemistry I. These subjects were chosen because they differ in the type of knowledge and response formats they elicit. Biology and nature and chemistry items typically require descriptive or concept-based answers, often expressed in short phrases or full sentences, whereas economics items more frequently involve reasoning process, definitions, or calculations.

```{r}
#| echo: true
exam_data <- exam_data %>%
  filter(vaknaam == "biologie" |
         vaknaam == "economie" |
         vaknaam == "natuur- en scheikunde I")
```

## Cleaning

```{r}
#| echo: true
# Remove HTML formatting first (before preprocessing)
exam_clean <- exam_data %>%
  mutate(
    # Remove HTML tags and entities
    response_nohtml = textclean::replace_html(response_clean),
    response_nohtml = str_squish(response_nohtml)
  )

# Rescore items to 0/1
exam_clean <- exam_clean %>%
  mutate(score = if_else(score == 2, 1, score))
```

## Integrate word length into the data

```{r}
# -----------------------------
# 1. Compute lengths
# -----------------------------

exam_clean <- exam_clean %>%
  mutate(
    char_len = nchar(response_nohtml),
    word_len = str_count(response_nohtml, "\\S+")
  )

# -----------------------------
# 2. Item-level length metrics
# -----------------------------

exam_item_length_raw <- exam_clean %>%
  group_by(item_label, vaknaam) %>%
  summarise(
    median_char_length = median(char_len, na.rm = TRUE),
    n_item_responses = n(),
    .groups = "drop"
  ) %>%
  group_by(vaknaam) %>%
  mutate(
    length_metric = (median_char_length - min(median_char_length, na.rm = TRUE)) /
      (max(median_char_length, na.rm = TRUE) - min(median_char_length, na.rm = TRUE) + 1e-9),
    item_length_group = ntile(length_metric, 3)
  ) %>%
  ungroup() %>%
  mutate(
    item_length_group = factor(
      item_length_group,
      levels = c(1, 2, 3),
      labels = c("Short", "Medium", "Long")
    )
  )

# -----------------------------
# 3. Clean join (no .x/.y EVER)
# -----------------------------

exam_clean <- exam_clean %>%
  select(
    -any_of(c(
      "median_char_length",
      "n_item_responses",
      "length_metric",
      "item_length_group"
    ))
  ) %>%
  left_join(
    exam_item_length_raw %>%
      select(
        item_label,
        vaknaam,
        median_char_length,
        n_item_responses,
        length_metric,
        item_length_group
      ),
    by = c("item_label", "vaknaam")
  )

# -----------------------------
# 4. Descriptives
# -----------------------------

## TABLE 1: Median word length per item length group and subject
length_descriptives <- exam_clean %>%
  group_by(vaknaam, item_length_group) %>%
  summarise(
    median_words = median(word_len, na.rm = TRUE),
    mean_words = mean(word_len, na.rm = TRUE),
    min_words = min(word_len, na.rm = TRUE),
    max_words = max(word_len, na.rm = TRUE),
    n_responses = n(),
    n_items = n_distinct(item_label),
    .groups = "drop"
  )

length_descriptives
```

## Pre-process

```{r}
# Initialize spaCy once before running the iterative procedure
spacy_initialize(
  model = "nl_core_news_sm",
  entity = FALSE
)

# Lowercase preprocessing function
lowercase_step <- function(x) {
  str_to_lower(x)
}

# Punctuation removal preprocessing function
punctuation_step <- function(x) {
  str_squish(str_remove_all(x, "[[:punct:]]"))
}

# Stop-words removal preprocessing function
stopword_step <- function(x) {
  keep_words <- c("ja", "nee", "niet")
  sw_nl <- setdiff(stopwords("nl"), keep_words)
  sw_nl_esc <- str_replace_all(sw_nl, "([\\W])", "\\\\\\1")

  str_squish(
    str_remove_all(
      x,
      paste0("\\b(", paste(sw_nl_esc, collapse = "|"), ")\\b")
    )
  )
}

# Lemmatization removal preprocessing function
lemmatize_spacy_unique <- function(x, batch_size = 10000) {
  x <- coalesce(x, "")
  
  lookup <- tibble(
    response_in = unique(x)
  ) %>%
    mutate(doc_id = as.character(row_number()))
  
  split_index <- split(
    seq_len(nrow(lookup)),
    ceiling(seq_len(nrow(lookup)) / batch_size)
  )
  
  out_list <- vector("list", length(split_index))
  
  # This processes only unique intermediate responses and optionally in batches,
  # which reduces redundant spaCy calls while preserving identical outputs.
  for (i in seq_along(split_index)) {
    idx <- split_index[[i]]
    batch_lookup <- lookup[idx, ]
    
    texts <- batch_lookup$response_in
    names(texts) <- batch_lookup$doc_id
    
    parsed <- spacy_parse(
      texts,
      lemma = TRUE,
      pos = FALSE,
      tag = FALSE,
      entity = FALSE,
      dependency = FALSE,
      nounphrase = FALSE,
      multithread = FALSE
    )
    
    batch_lemma <- parsed %>%
      group_by(doc_id) %>%
      summarise(
        response_out = str_squish(paste(lemma, collapse = " ")),
        .groups = "drop"
      )
    
    out_list[[i]] <- batch_lookup %>%
      left_join(batch_lemma, by = "doc_id") %>%
      mutate(response_out = coalesce(response_out, response_in)) %>%
      select(response_in, response_out)
  }
  
  bind_rows(out_list)
}

lemmatization_step <- function(x, batch_size = 10000) {
  lookup <- lemmatize_spacy_unique(x, batch_size = batch_size)
  lookup$response_out[match(x, lookup$response_in)]
}

apply_named_step <- function(x, step_name, batch_size = 10000) {
  if (step_name == "lowercase") {
    lowercase_step(x)
  } else if (step_name == "punctuation") {
    punctuation_step(x)
  } else if (step_name == "lemmatization") {
    lemmatization_step(x, batch_size = batch_size)
  } else if (step_name == "stopword") {
    stopword_step(x)
  } else {
    stop("Unknown preprocessing step: ", step_name)
  }
}

# ============================================================
# ITERATIVE PREPROCESSING WITH SCORE RESOLUTION
# ============================================================

resolve_score <- function(mean_score) {
  dplyr::case_when(
    is.na(mean_score) ~ NA_real_,
    mean_score < 0.40 ~ 0,
    mean_score <= 0.60 ~ 0.5,
    mean_score > 0.60 ~ 1
  )
}

make_size_bin <- function(n) {
  dplyr::case_when(
    n == 1 ~ "1",
    n >= 2  & n <= 3  ~ "2-3",
    n >= 4  & n <= 5  ~ "4-5",
    n >= 6  & n <= 15 ~ "6-15",
    n >= 16 & n <= 30 ~ "16-30",
    n > 30 ~ ">30",
  )
}

summarise_stage_resolved <- function(data, stage_name) {
  data %>%
    group_by(vaknaam, item_label, item_length_group, response_current) %>%
    summarise(
      n = n(),
      mean_score = mean(score_current, na.rm = TRUE),
      sd_score = sd(score_current, na.rm = TRUE),
      n_score_values = n_distinct(score_current[!is.na(score_current)]),
      inconsistent = n_score_values > 1,
      .groups = "drop"
    ) %>%
    mutate(
      stage = stage_name,
      resolved_score = resolve_score(mean_score),
      size_bin = make_size_bin(n)
    )
}

run_iterative_preprocessing_resolved <- function(
    data,
    step_order = c("lowercase", "punctuation", "lemmatization", "stopword"),
    batch_size = 5000
) {
  
  step_labels <- c(
    lowercase = "Lowercasing",
    punctuation = "No Punctuation",
    lemmatization = "Lemmatization",
    stopword = "Stopword Removal"
  )
  
  working_data <- data %>%
    mutate(row_id = row_number()) %>%
    transmute(
      row_id,
      vaknaam,
      item_label,
      item_length_group,
      response_current = coalesce(response_nohtml, ""),
      score_current = score
    )
  
  stage_summaries <- list()
  transition_summaries <- list()
  text_list <- list()
  
  # ----------------------------
  # Stage 0: cleaned responses
  # ----------------------------
  stage_summaries[[1]] <- summarise_stage_resolved(
    working_data,
    stage_name = "0. Cleaned"
  )
  
  text_list[[1]] <- working_data$response_current
names(text_list)[1] <- "0. Cleaned"

  # Resolve scores after Stage 0
  working_data <- working_data %>%
    left_join(
      stage_summaries[[1]] %>%
        select(
          vaknaam,
          item_label,
          item_length_group,
          response_current,
          resolved_score
        ),
      by = c("vaknaam", "item_label", "item_length_group", "response_current")
    ) %>%
    mutate(score_current = resolved_score) %>%
    select(-resolved_score)
  
  # ----------------------------
  # Iterative preprocessing steps
  # ----------------------------
  for (i in seq_along(step_order)) {
    
    step_name <- step_order[i]
    stage_name <- paste0(i, ". ", step_labels[[step_name]])
    
    # clusters before applying the step
    source_clusters <- working_data %>%
      group_by(
        vaknaam,
        item_label,
        item_length_group,
        response_before = response_current
      ) %>%
      summarise(
        n_before = n(),
        from_bin = make_size_bin(n_before),
        source_score = mean(score_current, na.rm = TRUE),
        .groups = "drop"
      )
    
    # apply preprocessing step to individual responses
    working_data <- working_data %>%
      mutate(
        response_before = response_current,
        response_current = apply_named_step(
          response_current,
          step_name = step_name,
          batch_size = batch_size
        )
      )
    
    text_list[[i + 1]] <- working_data$response_current
names(text_list)[i + 1] <- stage_name

    # summarise clusters after this step
    stage_summary <- summarise_stage_resolved(
      working_data,
      stage_name = stage_name
    )
    
    stage_summaries[[i + 1]] <- stage_summary
    
    # track movement from old cluster bins to new cluster bins
    transition_summaries[[i]] <- tibble()
    
    # resolve scores before going to the next preprocessing step
    working_data <- working_data %>%
      left_join(
        stage_summary %>%
          select(
            vaknaam,
            item_label,
            item_length_group,
            response_current,
            resolved_score
          ),
        by = c("vaknaam", "item_label", "item_length_group", "response_current")
      ) %>%
      mutate(score_current = resolved_score) %>%
      select(-resolved_score, -response_before)
  }
  
  processed_data <- data %>%
  mutate(!!!setNames(text_list, names(text_list)))

  list(
    processed_data = processed_data,
    stage_summary = bind_rows(stage_summaries),
    transitions = bind_rows(transition_summaries)
  )

}

# Define the default preprocessing order used in the main analysis
default_order <- c("lowercase", "punctuation", "lemmatization", "stopword")

# Not used, but put as an example, if you were to experiment with other ordering
alternative_order <- c("stopword", "lemmatization", "lowercase", "punctuation")

# Run the iterative preprocessing pipeline using the default stage order, feel free to experiment
result_resolved <- run_iterative_preprocessing_resolved(
  data = exam_clean,
  step_order = default_order,
  batch_size = 10000
)

# Extract the cluster-level summary across all preprocessing stages
stage_summary <- result_resolved$stage_summary

# Extract the response-level dataset with processed text columns added
exam_processed <- result_resolved$processed_data
```

## Save Preprocessed Data

```{r}
#| include: false
save(exam_processed, file = "../Data/02_preprocessed/data_preprocessed.RData")
save(exam_processed, file = "../Output/01_preprocessed/data_preprocessed.RData")
save(stage_summary, file = "../Output/02_results/stage_summary.RData")
```
