---
title: "Text Preprocessing"
format: revealjs
editor: visual
---

## Load Libraries

```{r}
#| include: false
# Load packages
library(dplyr)
library(stringr)
library(textclean)
library(stringdist)
library(stopwords)
library(spacyr)
library(tidyr)
library(purrr)
library(ggplot2)
library(ggalluvial)
library(xtable)
```

## Load Preprocessed Data

```{r}
#| include: false
load("../Output/01_preprocessed/data_preprocessed.RData")
load("../Output/02_results/stage_summary.RData")
```

## Selected Items

To illustrate how preprocessing affects both the structure of response clusters and the occurrence of these inconsistencies, a subset of nine items was selected based on the presence of sufficient non-singleton clusters and observable inconsistencies at baseline. One item is selected for each item length group and each subject to ensure coverage across conditions. These items are the public ones.

```{r}
selected_item_labels <- c("32f65m", "32cduw", "32ddak", "32jwg8", "32eq6b", "32ggsx", "32eyee", "328hr6", "32la75")
```

## Results

### Table 2: Distribution of cluster sizes and inconsistencies across preprocessing stages

```{r}
# Adjusted item labels and order
item_lookup <- tibble::tribble(
  ~Item,          ~item_label,
  "Item\\_1a",   "32f65m",
  "Item\\_1b",   "32cduw",
  "Item\\_1c",   "32ddak",
  "Item\\_2a",   "32jwg8",
  "Item\\_2b",   "32eq6b",
  "Item\\_2c",   "32ggsx",
  "Item\\_3a",   "32eyee",
  "Item\\_3b",   "328hr6",
  "Item\\_3c",   "32la75"
)

# Assign each cluster to a size bin based on the number of responses
make_bin <- function(n) {
  dplyr::case_when(
    n == 1 ~ "1",
    n >= 2  & n <= 3  ~ "2-3",
    n >= 4  & n <= 5  ~ "4-5",
    n >= 6  & n <= 15 ~ "6-15",
    n >= 16 & n <= 30 ~ "16-30",
    n > 30 ~ ">30",
  )
}

# Define the display order of the cluster-size bins
bin_levels <- c("1", "2-3", "4-5", "6-15", "16-30", ">30")

# Create the summary table of total and inconsistent clusters by item, stage, and size bin
bin_table_all_items <- stage_summary %>%
  semi_join(item_lookup, by = "item_label") %>%
  left_join(item_lookup, by = "item_label") %>%
  mutate(
    Item = factor(Item, levels = item_lookup$Item),
    size_bin = factor(make_bin(n), levels = bin_levels),
    stage = factor(
      stage,
      levels = c(
        "0. Cleaned",
        "1. Lowercasing",
        "2. No Punctuation",
        "3. Lemmatization",
        "4. Stopword Removal"
      )
    )
  ) %>%
  group_by(Item, stage, size_bin) %>%
  summarise(
    total_groups = dplyr::n(),
    inconsistent_groups = sum(inconsistent, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    cell = ifelse(
      size_bin == "1",
      as.character(total_groups),
      paste0(total_groups, " (", inconsistent_groups, ")")
    )
  ) %>%
  select(Item, stage, size_bin, cell) %>%
  pivot_wider(
    names_from = size_bin,
    values_from = cell,
    values_fill = "0 (0)"
  ) %>%
  arrange(Item, stage) %>%
  rename(
    Stage = stage,
    `$>$30` = `>30`
  )

# Display the final wide-format table in R
bin_table_all_items

# Blank repeated item labels, showing each item only once
bin_table_export <- bin_table_all_items %>%
  group_by(Item) %>%
  mutate(
    Item = ifelse(row_number() == ceiling(n() / 2), as.character(Item), "")
  ) %>%
  ungroup()

# Mark first row of each item block
bin_sep <- bin_table_all_items %>%
  arrange(Item, Stage) %>%
  group_by(Item) %>%
  mutate(
    new_group = row_number() == 1
  ) %>%
  ungroup()

# Store the row positions where item block separators should be inserted
add_lines <- which(bin_sep$new_group)

# Convert the table to a LaTeX longtable objec
xt <- xtable(
  bin_table_export,
  caption = "Distribution of cluster sizes and inconsistencies across preprocessing stages",
  label = "tab:bin_table"
)

# Export the formatted LaTeX table
print(
  xt,
  file = "bin_table_all_items.tex",
  include.rownames = FALSE,
  tabular.environment = "longtable",
  floating = FALSE,
  sanitize.text.function = identity,
  add.to.row = list(
    pos = as.list(add_lines - 1),
    command = rep("\\hline \n", length(add_lines))
  )
)
```

### Figure 3: Consistent and Inconsistent Cluster Sizes

```{r}
plot_bins_line <- stage_summary %>%
  semi_join(tibble(item_label = selected_item_labels), by = "item_label") %>%
  filter(size_bin != "1") %>%  # keep n = 1 in data, exclude from plot
  mutate(
    size_bin = factor(make_bin(n), levels = bin_levels),
    stage = factor(
      stage,
      levels = c(
        "0. Cleaned",
        "1. Lowercasing",
        "2. No Punctuation",
        "3. Lemmatization",
        "4. Stopword Removal"
      )
    )
  ) %>%
  group_by(vaknaam, item_label, item_length_group, stage, size_bin) %>%
  summarise(
    total_clusters = n(),
    inconsistent_clusters = sum(inconsistent, na.rm = TRUE),
    prop_inconsistent = inconsistent_clusters / total_clusters,
    .groups = "drop"
  ) %>%
  ggplot(
    aes(
      x = stage,
      y = inconsistent_clusters,
      color = size_bin,
      group = size_bin
    )
  ) +
  geom_line(linewidth = 0.8) +
  geom_point(aes(size = total_clusters), alpha = 0.8) +
  facet_grid(item_length_group ~ vaknaam, scales = "free_y") +
  labs(
    x = "Preprocessing stage",
    y = "Number of inconsistent clusters",
    color = "Cluster size bin",
    size = "Total clusters",
    caption = "Singleton clusters are excluded from the plot for readability."
  ) +
  scale_size_continuous(breaks = c(20, 50, 100, 150)) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

plot_bins_line
ggsave("../Output/02_results/stacked-lines.png",
       plot = plot_bins_line,
       width = 8,
       height = 6,
       dpi = 300)
```

### Figure 4: Distribution of Average Scores for Inconsistent Clusters

```{r}
plot_score_distribution <- stage_summary %>%
  semi_join(tibble(item_label = selected_item_labels), by = "item_label") %>%
  filter(size_bin != "1", inconsistent) %>%
  mutate(
    size_bin = factor(make_bin(n), levels = bin_levels)
  ) %>%
  ggplot(aes(x = mean_score)) +
  geom_histogram(binwidth = 0.1, boundary = 0, closed = "left") +
  facet_grid(size_bin ~ stage) +
  labs(
    x = "Average score of inconsistent cluster",
    y = "Number of inconsistent clusters"
  ) +
  theme_minimal(base_size = 11)

plot_score_distribution

ggsave("../Output/02_results/score-distribution.png",
       plot = plot_score_distribution,
       width = 8,
       height = 6,
       dpi = 300)
```

### Tables 3-6: Transition Matrices

```{r}
# -----------------------------
# 1. Define preprocessing stages
# -----------------------------

stage_cols <- c(
  "0. Cleaned",
  "1. Lowercasing",
  "2. No Punctuation",
  "3. Lemmatization",
  "4. Stopword Removal"
)

cluster_cols <- c(
  "cluster_0_Cleaned",
  "cluster_1_Lowercasing",
  "cluster_2_No_Punctuation",
  "cluster_3_Lemmatization",
  "cluster_4_Stopword_Removal"
)

score_col <- "score"

bin_levels <- c("1", "2-3", "4-5", "6-15", "16-30", ">30")

make_bin <- function(n) {
  case_when(
    n == 1 ~ "1",
    n >= 2  & n <= 3  ~ "2-3",
    n >= 4  & n <= 5  ~ "4-5",
    n >= 6  & n <= 15 ~ "6-15",
    n >= 16 & n <= 30 ~ "16-30",
    n > 30 ~ ">30",
    TRUE ~ NA_character_
  )
}


# -----------------------------
# 2. Create cluster IDs per stage
# -----------------------------

exam_processed_clustered <- exam_processed

for (s in stage_cols) {
  new_col <- paste0("cluster_", gsub("[^A-Za-z0-9]+", "_", s))

  exam_processed_clustered <- exam_processed_clustered %>%
    group_by(item_label, .data[[s]]) %>%
    mutate("{new_col}" := cur_group_id()) %>%
    ungroup()
}


# -----------------------------
# 3. Create inconsistency indicator per cluster
# -----------------------------

for (cl in cluster_cols) {
  new_col <- paste0("inconsistent_", cl)

  exam_processed_clustered <- exam_processed_clustered %>%
    group_by(item_label, .data[[cl]]) %>%
    mutate("{new_col}" := n_distinct(.data[[score_col]], na.rm = TRUE) > 1) %>%
    ungroup()
}


# -----------------------------
# 4. Bin function
# -----------------------------

make_bin <- function(n) {
  case_when(
    n == 1 ~ "1",
    n >= 2  & n <= 3  ~ "2-3",
    n >= 4  & n <= 5  ~ "4-5",
    n >= 6  & n <= 15 ~ "6-15",
    n >= 16 & n <= 30 ~ "16-30",
    n > 30 ~ ">30",
    TRUE ~ NA_character_
  )
}


# -----------------------------
# 5. Transition matrix function
# -----------------------------

make_transition_matrix <- function(data,
                                   from_cluster,
                                   to_cluster,
                                   inconsistency_col) {

  data %>%
    group_by(item_label, .data[[from_cluster]]) %>%
    mutate(from_size = n()) %>%
    ungroup() %>%
    group_by(item_label, .data[[to_cluster]]) %>%
    mutate(to_size = n()) %>%
    ungroup() %>%
    mutate(
      from_bin = factor(make_bin(from_size), levels = bin_levels),
      to_bin   = factor(make_bin(to_size), levels = bin_levels)
    ) %>%
    distinct(
      item_label,
      from_cluster_id = .data[[from_cluster]],
      to_cluster_id   = .data[[to_cluster]],
      from_bin,
      to_bin,
      inconsistent_after = .data[[inconsistency_col]]
    ) %>%
    group_by(from_bin, to_bin) %>%
    summarise(
      n_groups = n(),
      n_inconsistent = sum(inconsistent_after == TRUE, na.rm = TRUE),
      cell = paste0(n_groups, " (", n_inconsistent, ")"),
      .groups = "drop"
    ) %>%
    select(from_bin, to_bin, cell) %>%
    complete(
      from_bin = factor(bin_levels, levels = bin_levels),
      to_bin   = factor(bin_levels, levels = bin_levels),
      fill = list(cell = "0 (0)")
    ) %>%
    pivot_wider(
      names_from = to_bin,
      values_from = cell
    ) %>%
    arrange(factor(from_bin, levels = bin_levels)) %>%
    select(from_bin, `1`, `2-3`, `4-5`, `6-15`, `16-30`, `>30`)
}

# Table 1: Cleaned -> Lowercasing
tab_clean_lower <- make_transition_matrix(
  exam_processed_clustered,
  from_cluster = "cluster_0_Cleaned",
  to_cluster = "cluster_1_Lowercasing",
  inconsistency_col = "inconsistent_cluster_1_Lowercasing"
)

tab_clean_lower

# Table 2: Lowercasing -> No Punctuation
tab_lower_nopunct <- make_transition_matrix(
  exam_processed_clustered,
  from_cluster = "cluster_1_Lowercasing",
  to_cluster = "cluster_2_No_Punctuation",
  inconsistency_col = "inconsistent_cluster_2_No_Punctuation"
)

tab_lower_nopunct

# Table 3: No Punctuation -> Lemmatization
tab_nopunct_lemma <- make_transition_matrix(
  exam_processed_clustered,
  from_cluster = "cluster_2_No_Punctuation",
  to_cluster = "cluster_3_Lemmatization",
  inconsistency_col = "inconsistent_cluster_3_Lemmatization"
)

tab_nopunct_lemma

# Table 4: Lemmatization -> Stopword Removal
tab_lemma_stopword <- make_transition_matrix(
  exam_processed_clustered,
  from_cluster = "cluster_3_Lemmatization",
  to_cluster = "cluster_4_Stopword_Removal",
  inconsistency_col = "inconsistent_cluster_4_Stopword_Removal"
)

tab_lemma_stopword

# export
print(
  xtable(tab_clean_lower,
         caption = "Transition matrix from cleaned to lowercased responses.",
         label = "tab:tab-0-1"),
  include.rownames = FALSE,
  floating = TRUE,
  table.placement = "H",
  caption.placement = "top",
  file = "../Output/02_results/tab_clean_lower.tex"
)

print(
  xtable(tab_lower_nopunct,
         caption = "Transition matrix from lowercased to no-punctuation responses.",
         label = "tab:tab-1-2"),
  include.rownames = FALSE,
  floating = TRUE,
  table.placement = "H",
  caption.placement = "top",
  file = "../Output/02_results/tab_lower_nopunct.tex"
)

print(
  xtable(tab_nopunct_lemma,
         caption = "Transition matrix from no-punctuation to lemmatized responses.",
         label = "tab:tab-2-3"),
  include.rownames = FALSE,
  floating = TRUE,
  table.placement = "H",
  caption.placement = "top",
  file = "../Output/02_results/tab_nopunct_lemma.tex"
)

print(
  xtable(tab_lemma_stopword,
         caption = "Transition matrix from lemmatized to stopword-removed responses.",
         label = "tab:tab-3-4"),
  include.rownames = FALSE,
  floating = TRUE,
  table.placement = "H",
  caption.placement = "top",
  file = "../Output/02_results/tab_lemma_stopword.tex"
)
```
